---
title: "Graphs of Opioid Deaths"
author: "Julia Thompson"
date: "11/30/2019"
output: html_document
---

## ER visits by county & distance to treatment facility

Is there any relationship between number of ER visits and distance to treatment facility?

```{r setup, include=FALSE}
library(tidyverse)
library(viridis)
library(arcos)
library(readxl)
library(plotly)
knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

```{r load datasets, include = FALSE}
opioid_death_data = janitor::clean_names(read_csv('./opioid_related_deaths.csv'))

county_pop = janitor::clean_names(read_excel("./data/county_pop.xlsx"))

opioid_treatment_distance = janitor::clean_names(read_csv('./distance_to_treatment.csv'))
```

```{r, include = FALSE}
# clean opioid death data #

opioid_death = opioid_death_data %>% 
  filter(year == 2017) %>% 
  select(-year)

county_pop = 
  county_pop %>% 
  mutate(
    county = str_replace(county, " County", "")
  )

opioid_treatment_dist = 
  opioid_treatment_distance %>% 
  rename(distance = value) %>% 
  filter(state == 'New York') %>%
  select(county, distance)

# get rid of "county" in opioid_treatment_distance

opioid_treatment_dist = 
  opioid_treatment_dist %>% 
  mutate(
    county = str_replace(county, " County", "")
  )

treatment_death = 
  left_join(opioid_treatment_dist, opioid_death, by = "county")

treatment_death_final = 
  left_join(treatment_death, county_pop, by = "county") %>% 
  select(-rank) %>% 
  mutate(
    deaths_rate = (opioid_poisoning_deaths / population)*100
  )
```

any relationship between rate and distance? Nope... the shorter the distance, the more deaths. Looking at rate, there is no real pattern? Kinda interesting that there's no relationship I guess.

```{r}
ggplot(treatment_death_final, aes(x = distance, y = opioid_poisoning_deaths))+
  geom_point()

ggplot(treatment_death_final, aes(x = distance, y = deaths_rate))+
  geom_point()
```

```{r, include = FALSE}
take_back = read_csv("./data/take_back.csv") %>% 
  janitor::clean_names() %>% 
  select(county) %>% 
  group_by(county) %>% 
  summarize(
    takeback_facilities = n()
  )
```

compare number of pill take back locations by county

```{r, include = FALSE}
treatment_death_final = treatment_death_final %>% 
  left_join(take_back, treatment_death_final, by = "county")

treatment_death_final = treatment_death_final %>% 
  mutate(
    takeback_facilities = replace_na(takeback_facilities, 0)
  )
```

```{r}
ggplot(treatment_death_final, aes(x = distance, y = takeback_facilities))+
  geom_point()

ggplot(treatment_death_final, aes(x = takeback_facilities, y = deaths_rate))+
  geom_point()
```

https://www.health.ny.gov/statistics/opioid/data/pdf/nys_opioid_annual_report_2017.pdf

## Look at demographic data

```{r, include = FALSE}
deaths_age_grp = read_csv("./data/deaths_age_group.csv") %>% 
  janitor::clean_names() %>% 
  pivot_wider(
    names_from = age_group,
    values_from = opioid_poisoning_deaths
  ) %>% 
  janitor::clean_names()
```

Deaths by race by year:

```{r, include = FALSE}
deaths_by_year = deaths_age_grp %>% 
  select(year, region, race_or_ethnicity, sex, total) %>% 
  group_by(year, race_or_ethnicity) %>% 
  summarize(
    count = sum(total)
  )

ggplot(deaths_by_year, aes(x = year, y = count, fill = race_or_ethnicity)) +
  geom_bar(stat = "identity", position="dodge")
```

```{r}
deaths_by_year %>% 
  plot_ly(x = ~year, y = ~count, type = "bar",
          color = ~race_or_ethnicity, alpha = 0.5)
```

Deaths by age group by year

```{r, include = FALSE}
deaths_by_age = read_csv("./data/deaths_age_group.csv") %>% 
  janitor::clean_names() %>% 
  select(-region, -race_or_ethnicity, -sex) %>% 
  filter(age_group != "Total") %>% 
  group_by(year, age_group) %>% 
  summarize(
    count = sum(opioid_poisoning_deaths)
  )

ggplot(deaths_by_age, aes(x = year, y = count, fill = age_group)) +
  geom_bar(stat = "identity", position="dodge")

ggplot(deaths_by_age, aes(x = year, y = count, color = age_group)) +
  geom_point() +
  geom_line()

```

```{r}
deaths_by_age %>% 
    plot_ly(x = ~year, y = ~count, type = "bar",
          color = ~age_group, alpha = 0.5)

deaths_by_age %>% 
    plot_ly(x = ~year, y = ~count, type = "scatter", mode = "lines+markers",
          color = ~age_group, alpha = 0.5)
```


Might need to adjust for population, proportion of ethnicity, etc.

