---
title: "Statistical Learning"
output: 
  html_document:
    toc: true
    code_folding: hide
---

```{r setup, include=FALSE}
library(tidyverse)
library(viridis)
library(arcos)
library(plotly)
knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	message = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

```{r load datasets}
opioid_death_data = read_csv('./opioid_related_deaths.csv') %>% 
  janitor::clean_names()

opioid_er_data = read_csv('./opioid_related_visits.csv') %>% 
  janitor::clean_names()
                                        
opioid_treatment_distance = read_csv('./distance_to_treatment.csv') %>% 
  janitor::clean_names()

opioid_dem_df = read_csv('./data/deaths_age_group.csv') %>% 
  janitor::clean_names()
```

```{r pharma data}
prod_county = arcos::summarized_county_annual(state = "NY", key = "WaPo") %>% 
  janitor::clean_names()

county_pop = arcos::county_population(state = "NY", key = "WaPo") %>% 
  janitor::clean_names()
```


```{r data cleaning}

# clean opioid death data #

opioid_death_data = opioid_death_data %>%
  filter(year >= 2010) %>% 
  mutate(county = recode(county, "Kings (Brooklyn)" = "Kings",
                         "New York (Manhattan)" = "New York",
                         "St Lawrence" = "St. Lawrence"))

# clean opioid er data #

opioid_er_data = opioid_er_data %>% 
  select(year, patient_county_name, rural_urban, payer, er_opioid, inpatient_total_opioid, er_inpatient_total_opioid, outpatient_opioid, overall_opioid) %>%
  rename(county = patient_county_name)

# Combine Data Sets #

opioid_total_data = left_join(opioid_er_data, opioid_death_data, by = c('county', 'year')) %>% 
  arrange(county, year)
```

```{r pills per capita, results = "hide"}

# Pills bought by Pharmacies in Each county per Year #

pharma_df = left_join(prod_county, county_pop, by = c("buyer_county", "year")) %>% 
  select(county_name, year, count, population) %>% 
  rename(county = county_name,
         pills_bought = count) %>% 
  mutate(ppp = pills_bought/population)

```

## Spaghetti Plots

This exploratory analysis clusters opioid related data adjusted for population in each county 

by the intercepts and slopes. This involved joining our population data with opioid related

interventions in hospitals. The intercepts and slopes were extracted from linear models.

```{r opioid total by year, include = TRUE}
sum_df <- opioid_total_data %>% 
  group_by(county, year, rural_urban) %>% 
  summarize(er_opioid = sum(er_opioid),
            inpatient_total_opioid = sum(inpatient_total_opioid),
            er_inpatient_total_opioid = sum(er_inpatient_total_opioid),
            outpatient_opioid = sum(outpatient_opioid),
            overall_opioid = sum(overall_opioid),
            opioid_poisoning_deaths = sum(opioid_poisoning_deaths)
  )
```

```{r rural v urban, include = TRUE}
join_df <- inner_join(sum_df, pharma_df, by = c("county", "year"))
```

```{r, include = TRUE}
sum_total_df <- opioid_total_data %>% 
  group_by(county, year) %>% 
  summarize(er_opioid = sum(er_opioid),
            inpatient_total_opioid = sum(inpatient_total_opioid),
            er_inpatient_total_opioid = sum(er_inpatient_total_opioid),
            outpatient_opioid = sum(outpatient_opioid),
            overall_opioid = sum(overall_opioid),
            opioid_poisoning_deaths = sum(opioid_poisoning_deaths)
  )
```

```{r, include = TRUE}
death_sales_df <- inner_join(sum_total_df, pharma_df, by = c("county", "year"))
```


```{r, include = TRUE}
int_slope_dat <- pharma_df %>% 
  mutate(
    year = recode(year, 
                  "2006" = 0,
                  "2007" = 1,
                  "2008" = 2,
                  "2009" = 3,
                  "2010" = 4,
                  "2011" = 5,
                  "2012" = 6)
  ) %>% 
  group_by(county, year) %>% 
  summarize(ppp = pills_bought/population)
```

```{r}
int_slope_df <- int_slope_dat %>% 
  nest(data = year:ppp) %>%
  mutate(
    models = map(data, ~lm(ppp ~ year, data = .x)),
    result = map(models, broom::tidy)
  ) %>% 
  select(county, result) %>% 
  unnest(result) %>% 
  select(county, term, estimate) %>% 
  pivot_wider(
    names_from = term,
    values_from = estimate
  ) %>% 
  rename(int = "(Intercept)", slope = year)
```

```{r}
km_fit = 
  kmeans(
    int_slope_df[,c(2,3)]  %>% scale, 
    centers = 3)

int_slope_df =
  broom::augment(km_fit, int_slope_df)
```

```{r}
pills_plot <- left_join(int_slope_dat, int_slope_df) %>% 
  group_by(county) %>% 
  ggplot(aes(x = year, y = ppp, color = .cluster, text = county)) + 
  geom_point() + 
  geom_path() +
  scale_x_continuous(breaks = c(0, 1, 2, 3, 4, 5, 6), labels = c("2006", "2007", "2008", "2009", "2010", "2011", "2012")) +
  theme(legend.position = "none") +
  xlab("Year") +
  ylab("Pills Bought Per Person")
```

```{r, include = TRUE}
ggplotly(pills_plot, tooltip = "text")
```

```{r}
death_sales_dat <- death_sales_df %>%
  ungroup() %>% 
  mutate(
    year = recode(year, 
                  "2010" = 0,
                  "2011" = 1,
                  "2012" = 2)
  )
```

## Clustering with Opioid Deaths

```{r}
death_sales_dat <- death_sales_df %>% 
  select(county, year, opioid_poisoning_deaths, population) %>% 
  mutate(
    year = recode(year,
                  "2010" = 0,
                  "2011" = 1,
                  "2012" = 2)
  )
```

```{r}
death_int_slope_df <- death_sales_dat %>% 
  nest(data = year:population) %>%
  mutate(
    models = map(data, ~lm(opioid_poisoning_deaths/population*10000 ~ year, data = .x)),
    result = map(models, broom::tidy)
  ) %>% 
  select(county, result) %>% 
  unnest(result) %>% 
  select(county, term, estimate) %>% 
  pivot_wider(
    names_from = term,
    values_from = estimate
  ) %>% 
  rename(int = "(Intercept)", slope = year)
```

```{r}
death_km_fit = 
  kmeans(
    death_int_slope_df[,c(2,3)] %>% scale, 
    centers = 3)

death_int_slope_df =
  broom::augment(death_km_fit, death_int_slope_df)
```

```{r}
deaths_plot <- left_join(death_sales_dat, death_int_slope_df) %>% 
  group_by(county) %>% 
  ggplot(aes(x = year, y = opioid_poisoning_deaths/population*10000, color = .cluster, text = county)) + 
  geom_point() + 
  geom_path() +
  scale_x_continuous(breaks = c(0, 1, 2), labels = c("2010", "2011", "2012")) +
  theme(legend.position = "none") + 
  xlab("Year") +
  ylab("Opioid Related Deaths per 10000")
```

```{r, include = TRUE}
ggplotly(deaths_plot, tooltip = "text")
```

## Clustering with ER Visits

```{r}
er_sales_dat <- death_sales_df %>% 
  select(county, year, er_opioid, population) %>% 
  mutate(
    year = recode(year,
                  "2010" = 0,
                  "2011" = 1,
                  "2012" = 2)
  )
```

```{r}
er_int_slope_df <- er_sales_dat %>% 
  nest(data = year:population) %>%
  mutate(
    models = map(data, ~lm(er_opioid/population*10000 ~ year, data = .x)),
    result = map(models, broom::tidy)
  ) %>% 
  select(county, result) %>% 
  unnest(result) %>% 
  select(county, term, estimate) %>% 
  pivot_wider(
    names_from = term,
    values_from = estimate
  ) %>% 
  rename(int = "(Intercept)", slope = year)
```

```{r}
er_km_fit = 
  kmeans(
    er_int_slope_df[,c(2,3)] %>% scale, 
    centers = 3)

er_int_slope_df =
  broom::augment(er_km_fit, er_int_slope_df)
```

```{r}
er_plot <- left_join(er_sales_dat, er_int_slope_df) %>% 
  group_by(county) %>% 
  ggplot(aes(x = year, y = er_opioid/population*10000, color = .cluster, text = county)) + 
  geom_point() +
  geom_path() +
  scale_x_continuous(breaks = c(0, 1, 2), labels = c("2010", "2011", "2012")) +
  theme(legend.position = "none") +
  xlab("Year") +
  ylab("Opioid Related ER Visits per 10000")
```

```{r, include = TRUE}
ggplotly(er_plot, tooltip = "text")
```

## Scatter Plots

This scatter plot shows the number of opioid-related deaths per 10000 people versus the number of 

pills prescribed adjusted by population.

```{r, include = TRUE}
join_df %>% 
  ggplot(aes(x = ppp, y = opioid_poisoning_deaths/population*10000)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  xlab("Pills Bought per Person") +
  ylab("Opioid Related Deaths per 10000")
```

```{r}
fit <- lm(opioid_poisoning_deaths/population*10000 ~ ppp, data = join_df)
summary(fit)
```

## Model Building

For the purposes of this model I created a region variable that measures whether a county is in 

New York City or the rest of New York State. I tested two nested stepwise models to determine 

which model would be best at predicting the raw number of opioid poisoning deaths. Then, variables

in the larger model were removed starting with the highest VIF until no variables had a VIF score

greater than 10.

```{r}
join_dat <- join_df %>%
  ungroup() %>% 
  mutate(year = factor(year),
         region = ifelse(county %in% c("New York", "Kings", "Bronx", "Queens", "Richmond"), "NYC", "ROS")) %>% 
  select(-county)
fit1 <- lm(opioid_poisoning_deaths ~ . -pills_bought 
           -inpatient_total_opioid -er_inpatient_total_opioid -overall_opioid, data = join_dat)
fit2 <- lm(opioid_poisoning_deaths ~ region + pills_bought + rural_urban, data = join_dat)

model1 <- step(fit1)
model2 <- step(fit2)
```

```{r, include = TRUE}
summary(model1)
summary(model2)
```

```{r, include = TRUE}
anova(model1, model2)
```
 